One-arm spirals in locally isothermal, radially structured self-gravitating discs

### 2014-11-02 paper/discussion.tex
updated varq figs
minor updates

\section{Discussion}

@@

\subsection{Role of disc structure}

>>>>

\section{Discussion}
Having confirmed the $m=1$ spiral instability through numerical
simulations, we now use results from the simulations combined with
linear theory to construct an explanation for their growth. For
simplicitly, we return to 2D discs.

@@

\subsection{Role of self-gravity and disc structure} 
The background torque density $T_{BG}$ does not depend on
self-gravity. 
%and by definition, disc structure plays no role in the local description of linear modes 
However, our numerical simulations show that the $m=1$ spiral is
confined between two $Q$-barriers, where real solutions to the local
dispersion relation is possible (Eq. \ref{wavenumber}), as a result of the adopted disc
structure. Thus, the role of disc structure and self-gravity is to
allow a neutral mode to be set up and confined, which is then
destabilised through the background torque density.

%talk about additional simulations at high Q
%spiral mode as external potential to wider disc (lp11)

### 2014-11-02 paper/model.tex
minor updates

% We have purposefully adopted the simplest model that was found to
% develop one-arm spirals in discs with radial structure. 
The indirect potential associated with the non-inertial reference frame is 
neglected to avoid complications arising from the motion of the
central star. For the adopted disc masses $M_d\lesssim 0.1 M_*$, this
effect is not expected to be significant.

@@

The 2D disc equations are obtained by replacing $\rho$ with the surface
mass density $\Sigma$ and $p$ with the vertically-integrated pressure
$\Pi$ in the continuity and momenta equations, and the 2D fluid
velocity $\bm{v}=(v_R,v_\phi)$ is evaluated at the midplane, as are
the forces from the potential. In the Poisson equation, $\rho$ is
replaced by $\Sigma\delta(z)$, where $\delta(z)$ is the
delta function.

>>>>

% We have purposefully adopted the simplest model that was found to
% develop one-arm spirals in discs with radial structure. 
For the simulations presented in this paper, the indirect potential
associated with the non-inertial reference frame is purposefully 
neglected to avoid complications arising from the motion of the
central star. For the adopted disc masses $M_d\lesssim 0.1 M_*$, this
effect is not expected to be significant \citep{adams89}. Indeed,
simulations in the early stages of this project included the indirect
potential, and yield similar results.

%ang mom cons

@@

The 2D disc equations are obtained by replacing $\rho$ with the surface
mass density $\Sigma$, $p$ becomes the vertically-integrated pressure, 
%$\Pi$ in the continuity and momenta equations, 
and the 2D fluid
velocity $\bm{v}=(v_R,v_\phi)$ is evaluated at the midplane, as are
the forces from the potential. In the Poisson equation, $\rho$ is
replaced by $\Sigma\delta(z)$, where $\delta(z)$ is the
delta function.

### 2014-11-02 paper/paper.tex
added analytical discussion
minor updates

\maketitle
\begin{abstract} 
  We present numerical simulations of gravitational instability in
  astrophysical discs with an an imposed temperature profile, and in
  which only a sub-portion of it is self-gravitating. We conduct
  numerical hydrodynamic experiments in 2D using the FARGO code and in
  3D with the ZEUS-MP and PLUTO codes. In such systems we observe the
  development of a 
  spiral mode with azimuthal wavenumber $m=1$ with co-rotation radius well
  outside the bulk of the spiral arm, resulting in a nearly stationary
  one-arm spiral pattern in the self-gravitating region. This one-arm
  spiral does not require the movement of 
  the central star, as found in previous studies. Instead, we find it
  to be associated with a structured disc, as well as an imposed
  temperature gradient. We apply results
  from existing linear theory and attribute the growth of this spiral
  to the non-conservation of angular momentum for linear perturbations
  in discs with an imposed temperature gradient. This enables angular
  momentum exchange between the spiral and the background disc.

@@

\input{model}
\input{results}
%\input{results_3d}
%\input{discussion}
%\input{summary}

>>>>

\maketitle
\begin{abstract} 
  We present numerical simulations of gravitational instability in
  astrophysical discs with an an imposed temperature profile, and in
  which only a sub-portion of it is self-gravitating. We conduct
  numerical hydrodynamic experiments in 2D using the FARGO code and in
  3D with the ZEUS-MP and PLUTO codes. In such systems we observe the
  development of a 
  spiral mode with azimuthal wavenumber $m=1$ with co-rotation radius well
  outside the bulk of the spiral arm, resulting in a nearly stationary
  one-arm spiral pattern in the self-gravitating region. This one-arm
  spiral does not require the movement of 
  the central star, as found in previous studies. Instead, we find it
  to be associated with an imposed
  temperature gradient and the radial disc structure. 
  We apply results
  from existing linear theory and attribute the growth of this spiral
  to the non-conservation of angular momentum for linear perturbations
  in discs with a forced temperature gradient. This enables angular
  momentum exchange between the spiral and the background disc.

@@

\input{model}
\input{results}
\input{results_3d}
\input{discussion}
%\input{summary}

### 2014-11-02 paper/results.tex
minor updates

\subsection{Reference run with random perturbations}
In this fiducial simulation we subject the disc to initial perturbations in
cylindrical radial velocity, 
\begin{align}\label{randpert}
  v_R \to v_R+ c_s\frac{\delta}{M}
  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}_d}{\Delta 
          R_d}\right)^2\right]}\sum_{m=1}^M\cos{m\phi},
\end{align}
where $\delta$ is an amplitude, $\overline{R}_d = (R_{1}+R_{2})/2$
and $\Delta R_d = (R_{2}-R_{1})/2$. We use $M=10$ and choose 
$\delta\in[-10^{-3},10^{-3}]$ randomly but independent of $\phi$.

Fig. \ref{fargo_modeamp} plots evolution of the maximum
non-axisymmetric surface density amplitudes in $R\in[R_{1},R_{2}]$
for $m\in[1,10]$. Snapshots from the simulation are shown in
Fig. \ref{fargo_2d}. 
At early times $t\lesssim100P_0$ the disc 
dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
growth initially and saturate (or decays) after $t=40P_0$. Notice the
low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
possibily due to saturation of the high-$m$ modes  
\citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
again after $t=70P_0$, and eventually dominates the annulus.

@@

We plot growth rates of the $m=1$ spiral as a function of $q$ in 
Fig. \ref{fargo_varq_growth}. The correlation is well-fitted by a
linear relation
\begin{align*}
  \gamma = 0.014 q \Omega_k{R_0}. 
\end{align*}
This confirms that the temperature gradient is responsible for the
development of the one-arm spiral observed in our simulations. 
\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varq.ps}    
  \caption{Growth rates of the $m=1$ spiral mode as a function of the
    imposed sound-speed gradient $q$ (asterisks). A linear fit is also
    plotted (dotted line). 
    \label{fargo_varq_growth}} 
\end{figure}

>>>>

In these simulations we subject the disc to initial perturbations in 
cylindrical radial velocity, 
\begin{align}\label{randpert}
  v_R \to v_R+ c_s\frac{\delta}{M}
  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}}{\Delta 
        R}\right)^2\right]}\sum_{m=1}^M\cos{m\phi},
\end{align}
where the amplitude $\delta\in[-10^{-3},10^{-3}]$ is set randomly but
independent of $\phi$, $\overline{R} = (R_{1}+R_{2})/2$
and $\Delta R = (R_{2}-R_{1})/2$.

\subsection{Reference run with random perturbations}
To obtain a picture of the overall disc evolution, here we consider a
fiducial run initialised with $M=10$ in Eq. \ref{randpert}. 
%In this fiducial simulation we subject the disc to initial perturbations in
%cylindrical radial velocity, 
%\begin{align}\label{randpert}
%  v_R \to v_R+ c_s\frac{\delta}{M}
%  \exp{\left[-\frac{1}{2}\left(\frac{R-\overline{R}_d}{\Delta 
%          R_d}\right)^2\right]}\sum_{m=1}^M\cos{m\phi},
%\end{align}
%where $\delta$ is an amplitude, $\overline{R}_d = (R_{1}+R_{2})/2$
%and $\Delta R_d = (R_{2}-R_{1})/2$. We use $M=10$ and choose 
%$\delta\in[-10^{-3},10^{-3}]$ randomly but independent of $\phi$.   
Fig. \ref{fargo_modeamp} plots evolution of the maximum
non-axisymmetric surface density amplitudes in $R\in[R_{1},R_{2}]$
for $m\in[1,10]$. Snapshots from the simulation are shown in
Fig. \ref{fargo_2d}. 
At early times $t\lesssim100P_0$ the disc 
dominated by low-amplitude high-$m$ perturbations. The $m\geq4$ modes
growth initially and saturate (or decays) after $t=40P_0$. Notice the
low $m\leq 2$ modes decay initially, but grows between $t\in[20,40]P_0$,
possibily due to saturation of the high-$m$ modes  
\citep{laughlin96,laughlin97}. However, the $m=1$ mode begins to grow
again after $t=70P_0$, and eventually dominates the annulus.

@@

We plot growth rates of the $m=1$ spiral as a function of $q$ in 
Fig. \ref{fargo_varq_growth}. The correlation is well-fitted by a
linear relation
\begin{align*}
  \gamma \simeq 0.015 q \Omega_k(R_0). 
\end{align*}
This confirms that the temperature gradient is responsible for the
development of the one-arm spiral observed in our simulations. 
\begin{figure}
  \includegraphics[width=\linewidth]{figures/m1_analysis_plot_ratemax_fargo_varq.ps}    
  \caption{Growth rates of the $m=1$ spiral mode as a function of the
    imposed sound-speed gradient $q$ (asterisks). A linear fit is also
    plotted (dotted line). 
    \label{fargo_varq_growth}} 
\end{figure}

### 2014-11-02 paper/results_3d.tex
minor updates

\section{Three-dimensional simulations}
The 3D disc has radial size 
$[r_\mathrm{min},r_\mathrm{max}]=[0.4,10]R_0$ and vertical extent
$n_H=2$ scale-heights. The resolution is $N_r\times N_\theta\times
N_\phi=256\times32\times256$. Because of the much reduced resolution
compared to 2D, we use a smooth perturbation by setting
$\delta = 10^{-3}$ and $M=1$. This corresponds to a single $m=1$
spiral in the dead zone, which was found to dominate the 2D
simulation.

>>>>

\section{Three-dimensional simulations}
In this section we briefly review results from 3D simulations carried
out using the ZEUS-MP and PLUTO codes. The main purpose is to confirm
the physical reality of the

The 3D disc has radial size 
$[r_\mathrm{min},r_\mathrm{max}]=[0.4,10]R_0$ and vertical extent
$n_H=2$ scale-heights. The resolution is $N_r\times N_\theta\times
N_\phi=256\times32\times256$. Because of the much reduced resolution
compared to 2D, we use a smooth perturbation by setting
$\delta = 10^{-3}$ and $M=1$. This corresponds to a single $m=1$
spiral in the dead zone, which was found to dominate the 2D
simulation.
